### 2018-04-06 manuscript/main.tex
Add latex template for manuscript

%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{epstopdf}
% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{example/icml2018}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2018}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Bayesian prediction of intensity rates from multivariate point processes}

\begin{document}

\twocolumn[
\icmltitle{Probabilistic forecast of influenza from genetic and count data}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2018
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Aeiau Zzzz}{equal,to}
\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
\icmlauthor{Cieua Vvvvv}{goo}
\icmlauthor{Iaesut Saoeu}{ed}
\icmlauthor{Fiuea Rrrr}{to}
\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
\icmlauthor{Aaoeu Iasoh}{goo}
\icmlauthor{Buiui Eueu}{ed}
\icmlauthor{Aeuia Zzzz}{ed}
\icmlauthor{Bieea C.~Yyyy}{to,goo}
\icmlauthor{Teoau Xxxx}{ed}
\icmlauthor{Eee Pppp}{ed}
\end{icmlauthorlist}

\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}

\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract} 
\end{abstract}

\section{Introduction}

\section{Methods}

\section{Results}

\section{Discussion}

\bibliography{main}
\bibliographystyle{example/icml2018}

\end{document}

### 2018-04-06 R/correlation_HM450/sbatch_HNSC_FDR10.sh
Add batch script for HNSC and SKCm

@@ -3,7 +3,7 @@
 #SBATCH --job-name=corFDR10
 #SBATCH --output=log/corFDR10.%A.out
 #SBATCH --error=log/corFDR10.%A.err
-#SBATCH --time=0-2:00:00
+#SBATCH --time=0-1:00:00
 #SBATCH -p owners,normal
 #SBATCH --qos=normal
 #SBATCH --nodes=1
@@ -17,4 +17,4 @@ set -beEu -o pipefail
 
 module load R
 
-Rscript task_CGI_FDR10.R
+Rscript task_FDR10.R /oak/stanford/groups/andrewg/users/szmamie/repos/MethylationPrediction/data/HNSC/processed

### 2018-04-06 R/correlation_HM450/sbatch_SKCM_FDR10.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+#SBATCH --job-name=corFDR10
+#SBATCH --output=log/corFDR10.%A.out
+#SBATCH --error=log/corFDR10.%A.err
+#SBATCH --time=0-1:00:00
+#SBATCH -p owners,normal
+#SBATCH --qos=normal
+#SBATCH --nodes=1
+#SBATCH --cores=8
+#SBATCH --mem=64000
+#SBATCH --mail-type=END,FAIL
+#################
+# Usage: $ sbatch --array=1-2,4%1 sbatch.sh
+
+set -beEu -o pipefail
+
+module load R
+
+Rscript task_FDR10.R /oak/stanford/groups/andrewg/users/szmamie/repos/MethylationPrediction/data/SKCM/processed 

### 2018-04-06 R/correlation_HM450/task_FDR10.R

@@ -1,0 +1,33 @@
+# File name: task_FDR10.R
+# Author: Mamie Wang
+# Date created: 03/13/2018
+# Date modified: 03/13/2018
+# Input: processed RNAseq and methylation data
+# Output: methylation probes with correlation coefficients between matched pairs of RNAseq and methylation
+# probes on the subset of promoter + CGI/shore region probes less than FDR 0.1
+
+library(qvalue)
+source('computeCorrelation.R')
+
+args = commandArgs(trailingOnly = TRUE)
+if (length(args) == 0) {
+  stop('At least one arguments must be supplied', call.=F)
+}
+data.dir = args[1]
+
+load(paste0(data.dir, '/RNAseq.RData'))
+load(paste0(data.dir, '/HM450.RData'))
+
+q <- 0.1
+
+CGI.probes <- read.csv('../../data/promoterRegionProbes/CGI.tsv', 
+                            sep='\t', header=T, stringsAsFactors=F)
+probe.id <- as.character(CGI.probes[,1])
+correlation <- ComputeCorrelation(RNAseq, data.HM450, convert2M=T, subsetProbes=probe.id)
+p.value <- correlation$p.value
+q.obj <- qvalue(p = p.value, fdr.level=q)
+probes.significant <- matrix(rownames(correlation)[q.obj$significant], ncol=1)
+colnames(probes.significant) <- c('probe')
+write.table(probes.significant, file=paste0(data.dir, '/CGICorFdr10.tsv'), sep='\t', quote=F, row.names=F)
+print(dim(probes.significant)[1])
+

### 2018-04-06 R/regressionModel/task_subset.R

@@ -1,0 +1,45 @@
+#!/usr/bin/env Rscript
+# File name: task.R
+# Author: Mamie Wang
+# Date created: 03/07/2018
+# Date modified: 03/14/2018
+
+library(methods)
+source('regressionModel.R')
+
+args = commandArgs(trailingOnly = TRUE)
+if (length(args) == 0) {
+  stop('At least two arguments must be supplied', call.=F)
+}
+data.dir = args[1]
+
+load(paste0(data.dir, '/HM450.RData'))
+load(paste0(data.dir, '/RNAseq.RData'))
+
+CGI.probes <- read.csv(paste0(data.dir, '/CGICorFdr10.tsv'),
+                            sep='\t', header=T, stringsAsFactors=F)
+probe.id <- as.character(CGI.probes[,1])
+subsetProbes <- probe.id
+
+center <- T
+scale <- T
+convert2M <- T
+percent.test <- 0.3
+cutoff <- 20
+alpha <- 1
+distance <- 'euclidean'
+method <- 'ward.D2'
+imagefolder <- paste0(data.dir, '/moduleNetworkFigMcenterScale')
+datafolder <- paste0(data.dir, '/resultsMcenterScale')
+
+RunModel(data.HM450, RNAseq, imagefolder, datafolder, convert2M=convert2M, subsetProbes=subsetProbes, distance=distance, method=method, 
+         cutoff=cutoff, percent.test=percent.test, alpha=alpha, center=center, scale=scale)
+
+convert2M <- F
+imagefolder <- paste0(data.dir, '/moduleNetworkFigBcenterScale')
+datafolder <- paste0(data.dir, '/resultsBcenterScale')
+
+RunModel(data.HM450, RNAseq, imagefolder, datafolder, convert2M=convert2M, subsetProbes=subsetProbes, distance=distance, method=method, 
+         cutoff=cutoff, percent.test=percent.test, alpha=alpha, center=center, scale=scale)
+
+

### 2018-04-15 R/regressionModel/regressionModel.R
Compare clustering of common probes between HM450 and HM27

@@ -150,7 +150,7 @@ RunModel <- function(methylation, rnaseq, imagefolder, datafolder, convert2M=F,
   methylation <- processed$methylation
   rnaseq <- processed$rnaseq
   methylprobes.geneid <- methylation.processed$m.geneid[,2]
-  methylation <- AverageMethylation(methylation, methylprobes.geneid)
+  methylation <- AverageMethylation(methylation, methylprobes.geneid) # agregate the methylation level of probes of the same gene
   methylation <- data.matrix(methylation[,-1])
   rownames(methylation) <- methylation.processed$m.geneid[!duplicated(methylprobes.geneid), 1]
   

### 2018-04-18 notes/helperFunctions.R
Write the vbsr predictor information and module network plot

@@ -1,0 +1,94 @@
+# Helper functions for extract clustering information
+
+library(dplyr)
+library(biomaRt)
+library(fossil)
+
+ensembl <- useMart("ENSEMBL_MART_ENSEMBL", dataset="hsapiens_gene_ensembl", host="www.ensembl.org")
+
+
+#' Extract the probe id for given cluster and find corresponding gene name
+#'
+#' @param cluster the clustering membership for each probe
+#' @param k the cluster number
+#' @return A data frame containing the probe id and gene name in the given cluster
+ExtractClusterProbes <- function(cluster, geneid, k) {
+  probe.names <- names(cluster)[cluster==k]
+  probe.genes <- geneid %>%
+    dplyr::filter(ID %in% probe.names) %>%
+    dplyr::select(ID, Gene.Symbol)
+  # split fields with multiple gene symbol as separate rows
+  n.rows <- dim(probe.genes)[1]
+  ids <- c()
+  genes <- c()
+  for (i in seq(n.rows)) {
+      id <- probe.genes[i,  1]
+      gene <- probe.genes[i, 2]
+      gene <- strsplit(gene, ',')[[1]]
+      n.gene <- length(gene)
+      if (n.gene > 1) {
+          ids <- c(ids, rep(id, n.gene))
+          genes <- c(genes, gene)
+      } else {
+          ids <- c(ids, id)
+          genes <- c(genes, gene)
+      }
+  }
+  print(paste('The number of probes for cluster', k, 'is', n.rows, 'and the number of genes is', length(ids)))
+  return(data.frame(id=ids, gene=genes))
+}
+
+#' Merge the hgnc names with the hgnc description, NA if not found
+#'
+#' @param geneid a data frame containing the gene and probe id
+#' @return a data frame with gene description from hgnc if available
+MergeHGNCDescription <- function(geneid) {
+    genes.info <- getBM(c("hgnc_symbol", "description"), "hgnc_symbol", geneid$gene, mart=ensembl)
+    geneid %>%
+      left_join(genes.info, by=c('gene'='hgnc_symbol'))
+}
+
+#' Write csv file for data frame
+#'
+#' @param data the data frame to be written
+#' @param path the writing path
+#' @param None (side effect: write data frame to csv)
+WriteCSV <- function(data, path) {
+    write.table(data, file=path, sep=',', quote=F, 
+               row.names=F, col.names=T)
+    print(paste0('Wrote csv to ', path))
+}
+
+#' Wrapper function to extract cluster probes and merge description and write csv
+#'
[…]

### 2018-05-15 R/regressionModel/vbsr-run.sh
Update predictor frequency table

@@ -4,7 +4,7 @@
 #SBATCH --output=log/vbsr.%A_%a.out
 #SBATCH --error=log/vbsr.%A_%a.err
 #SBATCH --time=0-01:00:00
-#SBATCH -p normal
+#SBATCH -p andrewg
 #SBATCH --qos=normal
 #SBATCH --nodes=1
 #SBATCH --ntasks=4

### 2018-05-15 manuscript/main.tex
Add method section for the pipeline

@@
 \section{Methods}
+\begin{figure}[h]
+\includegraphics[width=0.5\textwidth]{../figs/flowChart}
+\caption{\textbf{Pipeline}}
+\end{figure}
+\subsection{data preprocessing}
+TCGA Level 3 mRNAseq and DNA methylation data (HumanMethylation450 arrays) was downloaded from Broad TCGA GDAC site for LAML, SKCM and HNSC.
+mRNAseq probes with missing values were removed.
+We queried the GO terms (molecular function and biological process) based on the HUGO symbol of the probes and kept only probes that contain keywords related to transcription and methylation as our feature matrix.
+
+Methylation array data was filtered to remove probes that are on chromosome X and Y and have missing values.
+Probes with UCSC RefGene group annotation as TSS 1500 and located within UCSC CpG island annotation were selected.
+The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with FDR adjusted p-value (FDR < 0.1) of 0.05.
+The resulting probes were clustered using hierarchical clustering with euclidean distance using Ward's method.
+20 clusters were set as the cutoff for assigning cluster membership.
+The average methylation level within the cluster of each patient was computed as the response vector.
+
+\subsection{variational Bayesian spike regression}
+variational Bayesian spike regression (VBSR) \cite{logsdon2012novel} is a Bayesian approximation of best subset selection.
+We used VBSR to select subset of informative mRNAseq probes that best explain the observed pattern of methylation level across patients out of all possible subset.
+VBSR approaches the problem by using a variational Bayes approximation that probabilistically models the inclusion of any probes in the regression model.
+The p-value on the estimated value of penalized regression coefficient were computed and significant probes were extracted at FDR adjusted p-value (FDR < 0.1) of 0.05.
+R package \texttt{vbsr} was used.
+
+\subsection{Geneset analysis of methylation clusters}
+The enrichment of biological processes in each methylation clusters were examined by comparing to pre-defined genesets on MSigDB by hypergeometric test.
 

### 2018-06-09 manuscript/main.tex
reproduce Figueroa clustering for 16 subtypes of patients
manuscrip method section on Figuoera data

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Bayesian prediction of intensity rates from multivariate point processes}
+\icmltitlerunning{??}
 
@@
 \twocolumn[
-\icmltitle{Probabilistic forecast of influenza from genetic and count data}
+\icmltitle{??}
 
@@
 \begin{icmlauthorlist}
-\icmlauthor{Aeiau Zzzz}{equal,to}
-\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-\icmlauthor{Cieua Vvvvv}{goo}
-\icmlauthor{Iaesut Saoeu}{ed}
-\icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
+\icmlauthor{Meng Wang}{bmi}
+\icmlauthor{Andrew J. Gentles}{bds}
 \end{icmlauthorlist}
 
-\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
-
-\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
+\icmlaffiliation{bmi}{Biomedical Informatics Training Program, Stanford University, CA, USA}
+\icmlaffiliation{bds}{Department of Biomedical Data Sciences, Stanford University, CA, USA}
+\icmlcorrespondingauthor{Andrew J. Gentles}{andrewg@stanford.edu}
 
@@
 
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 
 \begin{abstract}
+
 \end{abstract}
@@
 \section{Methods}
+
+\subsection{\citet{figueroa2010dna} DNA methylation and gene expression microarray data }
+
+DNA methylation and gene expression microarray data from \citet{figueroa2010dna} were downloaded from the GEO repository (Accession number ).
+We followed preprocessing procedure in \citet{figueroa2010dna} and performed unsupervised hierarchical clustering of patients from methylation microarray using the subset of probe set ($n = 3745$) with standard deviation $> 1$ across all AML cases.
+Lingoes transformed 1 - Pearson correlation distance is used as the distance metrics and Ward's method is used as the clustering algorithm to reproduce the same patient groupings from original study.
+Gene expression array \citep{precog...}
+
+\subsection{Consensus clustering of methylation probes}
+
+Consensus clustering \citep{monti2003consensus, wilkerson2010consensusclusterplus} was used determine the optimal number of clusters of probes. [parameters]
+
 \begin{figure}[h]
@@
 \end{figure}
-\subsection{data preprocessing}
+
+\subsection{TCGA data preprocessing}
 TCGA Level 3 mRNAseq and DNA methylation data (HumanMethylation450 arrays) was downloaded from Broad TCGA GDAC site for LAML, SKCM and HNSC.
@@
 Probes with UCSC RefGene group annotation as TSS 1500 and located within UCSC CpG island annotation were selected.
-The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with FDR adjusted p-value (FDR < 0.1) of 0.05.
+The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with FDR adjusted p-value (FDR $< 0.1$) of 0.05.
 The resulting probes were clustered using hierarchical clustering with euclidean distance using Ward's method.
@@
 R package \texttt{vbsr} was used.
+[parameters]
 
@@
 \section{Results}
+
+\subsection{Consensus clustering of methylation probes}
+
+Consensus clustering \citep{monti2003consensus, wilkerson2010consensusclusterplus} provides quantitative and visual stability evidence for number of groups present in the dataset and were used to cluster the methylation probes.
+Figure \ref{CC}.
+
+\begin{figure*}[ht!]
+\centering
+\includegraphics[width=\textwidth]{Figures/clustering}
+\caption{\textbf{Consensus clustering (CC) of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$.
+$k = 7$ shows high intra-cluster consensus and low intercluster consensus.
+(b) cumulative distribution function of consensus matrix at each $k = 1, \dots, 10$.
+$k = 7$ approaches the maximum consensus distribution (c) Area under CDF of consensus matrix for $k = 1, \dots, 10$.
+$k = 7$ is the largest $k$ with a appreciable increase in consensus (d) Hierarchical clustering of cases and probes.
+Each row represents a probe and each column represents a patients.
+Methylation intensity level were row and column normalized.
+The16 clusters of AML cases from \citet{figueroa2010dna} were reproduced.
+Probes were clustered using Ward's method with Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.
+}
+\label{CC}
+\end{figure*}
+
+\subsection{Gene enrichment analysis}
+[table here]
 

### 2018-06-12 manuscript/main.tex
Add details to the method section

@@
 % Use the following line for the initial blind version submitted for review:
-\usepackage{example/icml2018}
+\usepackage[accepted]{example/icml2018}
 
@@
 
-\subsection{\citet{figueroa2010dna} DNA methylation and gene expression microarray data }
+%\begin{figure}[h]
+%\includegraphics[width=0.5\textwidth]{../figs/flowChart}
+%\caption{\textbf{Pipeline}}
+%\end{figure}
 
-DNA methylation and gene expression microarray data from \citet{figueroa2010dna} were downloaded from the GEO repository (Accession number ).
-We followed preprocessing procedure in \citet{figueroa2010dna} and performed unsupervised hierarchical clustering of patients from methylation microarray using the subset of probe set ($n = 3745$) with standard deviation $> 1$ across all AML cases.
-Lingoes transformed 1 - Pearson correlation distance is used as the distance metrics and Ward's method is used as the clustering algorithm to reproduce the same patient groupings from original study.
-Gene expression array \citep{precog...}
+\subsection{\citet{figueroa2010dna} DNA methylation and gene expression data}
 
-\subsection{Consensus clustering of methylation probes}
+DNA methylation and gene expression microarray data of 344 AML cases from \citet{figueroa2010dna} were downloaded from the GEO repository (GEO Accession number: GSE18700, GSE14468).
+Briefly, methylation data were obtained from custom design human promoter microarray for HELP assay, and gene expression data was obtained from Affymetrix Human Genome U133 Plus 2.0 Array.
 
-Consensus clustering \citep{monti2003consensus, wilkerson2010consensusclusterplus} was used determine the optimal number of clusters of probes. [parameters]
+Preprocessed gene expression data were obtained from \citet{gentles2015prognostic}, which normalized the raw CEL files with Affymetrix MAS5 algorithm and $\log_2$ transformed.
+We queried the GO terms on molecular function and biological process based on the HUGO gene symbol of the probes and filtered for probes that contain keywords related to transcription and methylation, which serves as predictors of our model.
 
-\begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{../figs/flowChart}
-\caption{\textbf{Pipeline}}
-\end{figure}
+For methylation data analysis, we followed \citet{figueroa2010dna} and performed unsupervised hierarchical clustering of patients from methylation microarray using subset of the probes with standard deviation $> 1$ across all AML cases ($n = 3745$).
+Lingoes transformed 1 - Pearson correlation distance and Ward's minimum variance criterion were used.
+The patient clusters from the original study was reproduced.
+We clustered the methylation probes using consensus clustering \citep{monti2003consensus}, which produces visual and quantitative stability evidence to a given number of clusters $k$ and cluster assignments.
+R package \texttt{ConsensusClusterPlus} \citep{wilkerson2010consensusclusterplus} provides an implementation of the method and were used to determine the optimal number of clusters.
+We used the following options: 80 \% probe subsampling, Ward's criterion with Lingoes transformation on 1 - Pearson correlation distance, 50 replicates for each $k$  and maximum $k = 10$.
+We decides the number of optimal number of clusters based on the consensus matrix and area under CDF.
+The enrichment of biological processes represented by genes in each methylation cluster were examined using PANTHER Classification System \citep{mi2013large}.
 
-\subsection{TCGA data preprocessing}
-TCGA Level 3 mRNAseq and DNA methylation data (HumanMethylation450 arrays) was downloaded from Broad TCGA GDAC site for LAML, SKCM and HNSC.
-mRNAseq probes with missing values were removed.
-We queried the GO terms (molecular function and biological process) based on the HUGO symbol of the probes and kept only probes that contain keywords related to transcription and methylation as our feature matrix.
+\subsection{Selection of significant gene expression probes}
+Variational Bayesian spike regression (vBsr) \cite{logsdon2012novel} is a penalized Bayesian regression model that imposes sparsity constraint on the regression coefficients using a spike-and-slab prior and utilized mean-field approximation to achieve fast computation.
+In addition, the algorithm was ran multiple times with random initialization to identify multiple local maxima of lower bound and provides the option of Bayesian Model Averaging (BMA) over the identified models to produce an estimate to reduce the model uncertainty.
+vBsr also defines a test statistic $z_{vb}$ associated with each penalized coefficients that allow control over the family-wise error rate by tuning the penalty parameter $l_0$ such that $z_{vb}$ statistics are approximately $\mathcal{N}(0, 1)$ under the null hypothesis.
+We used vBsr to select the subset of gene expression probes that significantly associate with the observed pattern of methylation level across patients.
+We ran 100 random starts for each model and used BMA option.
+We tuned the penalty parameter $l_0$ such that a feature will have a posterior probability of 0.95 if it passes a Bonferroni correction in the multivariate model to control the Type I error rate.
+Gene expression probes that were significant for $z_{vb}$ at $\widehat{FDR} = 0.1$ were selected.
 
-Methylation array data was filtered to remove probes that are on chromosome X and Y and have missing values.
-Probes with UCSC RefGene group annotation as TSS 1500 and located within UCSC CpG island annotation were selected.
-The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with FDR adjusted p-value (FDR $< 0.1$) of 0.05.
-The resulting probes were clustered using hierarchical clustering with euclidean distance using Ward's method.
-20 clusters were set as the cutoff for assigning cluster membership.
-The average methylation level within the cluster of each patient was computed as the response vector.
-
-\subsection{variational Bayesian spike regression}
-variational Bayesian spike regression (VBSR) \cite{logsdon2012novel} is a Bayesian approximation of best subset selection.
-We used VBSR to select subset of informative mRNAseq probes that best explain the observed pattern of methylation level across patients out of all possible subset.
-VBSR approaches the problem by using a variational Bayes approximation that probabilistically models the inclusion of any probes in the regression model.
-The p-value on the estimated value of penalized regression coefficient were computed and significant probes were extracted at FDR adjusted p-value (FDR < 0.1) of 0.05.
-R package \texttt{vbsr} was used.
-[parameters]
-
-\subsection{Geneset analysis of methylation clusters}
-The enrichment of biological processes in each methylation clusters were examined by comparing to pre-defined genesets on MSigDB by hypergeometric test.
-
-\section{Results}
-
-\subsection{Consensus clustering of methylation probes}
-
-Consensus clustering \citep{monti2003consensus, wilkerson2010consensusclusterplus} provides quantitative and visual stability evidence for number of groups present in the dataset and were used to cluster the methylation probes.
-Figure \ref{CC}.
-
-\begin{figure*}[ht!]
+\begin{figure*}[htb!]
 \centering
 \includegraphics[width=\textwidth]{Figures/clustering}
-\caption{\textbf{Consensus clustering (CC) of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$.
+\caption{\textbf{Consensus clustering of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$.
 $k = 7$ shows high intra-cluster consensus and low intercluster consensus.
@@
 
+\subsection{TCGA LAML RNA-seq and DNA methylation data}
+TCGA Level 3 mRNAseq and DNA methylation data was downloaded from Broad TCGA GDAC site for LAML.
+mRNAseq probes with missing values were imputed by KNN using R packages \texttt{impute} \citep{hastie2001impute}.
+We queried the GO terms on molecular function and biological process based on the HUGO symbol of the probes and filtered for probes that contain keywords related to transcription and methylation as predictors for our model.
+
+Methylation array data was obtained from Illumina Infinium HumanMethylation450 BeadChip.
+Probes that are on chromosome X and Y were removed.
+Probes with UCSC RefGene group annotation as TSS 1500 and located within UCSC CpG island annotation were selected.
+The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with FDR adjusted p-value (FDR $< 0.1$) of 0.05.
+The resulting probes were clustered using hierarchical clustering with euclidean distance using Ward's method.
+20 clusters were set as the cutoff for assigning cluster membership.
+The average methylation level within the cluster of each patient was computed as the response vector.
+
+\section{Results}
+
+\subsection{Significant associated gene expression probes in \citet{figueroa2010dna}}
+
+Figure \ref{CC} showed results of consensus clustering on the methylation probes for \citet{figueroa2010dna}.
+
 \subsection{Gene enrichment analysis}

### 2018-06-12 manuscript/main.tex
Add table of enriched biological process

@@
 The patient clusters from the original study was reproduced.
-We clustered the methylation probes using consensus clustering \citep{monti2003consensus}, which produces visual and quantitative stability evidence to a given number of clusters $k$ and cluster assignments.
+Next, we clustered the probes with similar methylation pattern using consensus clustering \citep{monti2003consensus}, which produces visual and quantitative stability evidence to a given number of clusters $k$ and cluster assignments.
 R package \texttt{ConsensusClusterPlus} \citep{wilkerson2010consensusclusterplus} provides an implementation of the method and were used to determine the optimal number of clusters.
@@
 We decides the number of optimal number of clusters based on the consensus matrix and area under CDF.
-The enrichment of biological processes represented by genes in each methylation cluster were examined using PANTHER Classification System \citep{mi2013large}.
+The enrichment of biological processes represented by genes in each methylation cluster were examined using PANTHER overrepresentation test using Gene Ontology Biological process annotation data \citep{mi2013large} with $ FDR < 0.05$.
 
@@
 vBsr also defines a test statistic $z_{vb}$ associated with each penalized coefficients that allow control over the family-wise error rate by tuning the penalty parameter $l_0$ such that $z_{vb}$ statistics are approximately $\mathcal{N}(0, 1)$ under the null hypothesis.
-We used vBsr to select the subset of gene expression probes that significantly associate with the observed pattern of methylation level across patients.
+We used vBsr to select the subset of gene expression probes that significantly associate with the observed pattern of variation of methylation level across patients.
 We ran 100 random starts for each model and used BMA option.
@@
 The16 clusters of AML cases from \citet{figueroa2010dna} were reproduced.
-Probes were clustered using Ward's method with Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.
-}
+Probes were clustered using Ward's method with Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.}
 \label{CC}
@@
 
-Figure \ref{CC} showed results of consensus clustering on the methylation probes for \citet{figueroa2010dna}.
+Figure \ref{CC} shows the results of consensus clustering on probes and significantly enriched biological processes in each cluster is shown below in Table \ref{enriched}.
 
-\subsection{Gene enrichment analysis}
-[table here]
+\begin{table*}[ht!]
+\centering
+\begin{tabular}{lllllllll}
+  \toprule
+Enriched biological process & ref & overlap & expected & fold enrich & raw P & FDR \\ 
+  \midrule
+  cluster 1 (size 326) \\
+  \midrule
+  cell communication & 5693 &  97 & 63.04 & 1.54 & 1.8e-6 & 2.4e-3 \\ 
+  signaling & 5578 &  96 & 61.77 & 1.55 & 1.5e-6 & 2.3e-3 \\ 
+   \midrule
+   cluster 3 (size 671) \\
+   \midrule
+     cellular process & 15478 & 436 & 379.56 & 1.15 & 6.1e-9 & 9.5e-5 \\ 
+\midrule
+  cluster 5 (size 816)\\
+  \midrule
+  developmental process  & 5654 & 310 & 199.38 & 1.55 & 9.2e-18 & 1.4e-13 \\ 
+  anatomical structure development & 5299 & 293 & 186.86 & 1.57 & 5.6e-17 & 4.4e-13 \\ 
+  multicellular organism development & 4918 & 275 & 173.42 & 1.59 & 2.7e-16 & 1.1e-12 \\ 
+  system development & 4309 & 244 & 151.95 & 1.61 & 1.0e-14 & 2.6e-11 \\ 
+     \midrule
+   cluster 6 (size 656)\\
+  \midrule
+  developmental process & 5654 & 199 & 147.52 & 1.35 & 2.2e-6 & 5.7e-3 \\ 
+  cellular component organization or biogenesis & 5773 & 196 & 150.62 & 1.30 & 3.1e-5 & 2.7e-2 \\ 
+  cellular component organization & 5584 & 192 & 145.69 & 1.32 & 2.0e-5 & 2.1e-2 \\ 
+  anatomical structure development & 5299 & 190 & 138.25 & 1.37 & 1.3e-6 & 5.2e-3 \\ 
+  multicellular organism development & 4918 & 184 & 128.31 & 1.43 & 1.1e-7 & 1.6e-3 \\ 
+     \midrule
+     cluster 7 (size 461) \\
+     \midrule
+     negative regulation of biological process & 5129 & 132 & 89.21 & 1.48 & 8.2e-07 & 1.3e-2 \\ 
+	\bottomrule
+\end{tabular}
+\caption{\textbf{Enriched biological processes in \citet{figueroa2010dna} methylation probes cluster} }
+\label{enriched}
+\end{table*}
 

### 2018-06-12 R/regressionModel/regressionModel.R
Figuoera vbsr plots

@@ -107,11 +107,10 @@ ModuleHeatmap <- function(methylation, avemethyl, rnaseq, filename=NULL, center=
   rownames(methylation) <- NULL
   colnames(methylation) <- NULL
   rownames(rnaseq) <- NULL
-  rownames(avemethyl) <- NULL
+  names(avemethyl) <- NULL
   ht1 <- Heatmap(methylation, col=inferno(10), width=unit(2.5, 'inches'), name='methylation', cluster_row=T, 
                  column_title='methylation')
-  ha1 <- avemethyl
-  ht2 <- Heatmap(ha1, col=inferno(10), width=unit(0.1, 'inches'), name='Average methylation')
+  ht2 <- Heatmap(avemethyl, col=inferno(10), width=unit(0.1, 'inches'), name='average methylation', column_title="")
   if (center) {
     GEP <- scale(rnaseq, center=center, scale=scale)
     if(!is.null(coef)) GEP <- t(t(GEP) * coef)
@@ -120,13 +119,12 @@ ModuleHeatmap <- function(methylation, avemethyl, rnaseq, filename=NULL, center=
   }
   ht3 <- Heatmap(GEP, col=inferno(10), width=unit(0.4, 'inches'),  name='GEP', column_title='GEP', cluster_columns=T)
   if(!is.null(filename)) {
-  setEPS()
-  postscript(file=filename, width=13, height=10)
-    draw(ht1 + ht2 + ht3)
-  dev.off()
-  print(paste('Image saved as', filename))
+    pdf(file=filename, width=6, height=10)
+      draw(ht1 + ht2 + ht3)
+    dev.off()
+    print(paste('Image saved as', filename))
   } else {
-    draw(ht1 + ht2 + ht3)
+    draw(ht1+ht2+ht3)
   }
 }
 

### 2018-06-12 manuscript/main.tex
Create the figure on vbsr

@@
 \includegraphics[width=\textwidth]{Figures/clustering}
-\caption{\textbf{Consensus clustering of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$.
-$k = 7$ shows high intra-cluster consensus and low intercluster consensus.
-(b) cumulative distribution function of consensus matrix at each $k = 1, \dots, 10$.
-$k = 7$ approaches the maximum consensus distribution (c) Area under CDF of consensus matrix for $k = 1, \dots, 10$.
-$k = 7$ is the largest $k$ with a appreciable increase in consensus (d) Hierarchical clustering of cases and probes.
+\caption{\textbf{Consensus clustering of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$ (b) cumulative distribution function of consensus matrix at each $k = 1, \dots, 10$ (c) Area under CDF of consensus matrix for $k = 1, \dots, 10$ (d) Hierarchical clustering of cases and probes.
 Each row represents a probe and each column represents a patients.
 Methylation intensity level were row and column normalized.
-The16 clusters of AML cases from \citet{figueroa2010dna} were reproduced.
-Probes were clustered using Ward's method with Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.}
+The16 AML cases from \citet{figueroa2010dna} were reproduced.
+Probes were clustered using Ward's method with lingoes transformation of 1 - Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.}
 \label{CC}
@@
 
-Figure \ref{CC} shows the results of consensus clustering on probes and significantly enriched biological processes in each cluster is shown below in Table \ref{enriched}.
+Figure \ref{CC} shows the results of consensus clustering on probes. [Explain the choice of 7 clusters]
+ $k = 7$ shows high intra-cluster consensus and low intercluster consensus.
+$k = 7$ approaches the maximum consensus distribution 
+$k = 7$ is the largest $k$ with a appreciable increase in consensus 
+ and significantly enriched biological processes in each cluster is shown below in Table \ref{enriched}.
 
@@
 
+\subsection{vBsr}
+
+\begin{figure*}[htb!]
+\includegraphics[width=\textwidth]{Figures/vBsr}
+\caption{\textbf{Heatmap of methylation clusters, average methylation level and gene expression probes significantly associate with average methylation level}}
+\end{figure*}
+
 \section{Discussion}

### 2018-06-12 notes/helperFunctions.R
LaTeX table for annotation

@@ -45,12 +45,7 @@ ExtractClusterProbes <- function(cluster, geneid, k) {
 #'
 #' @param geneid a data frame containing the gene and probe id
 #' @return a data frame with gene description from hgnc if available
-MergeHGNCDescription <- function(geneid) {
-    genes.info <- getBM(c("hgnc_symbol", "description"), "hgnc_symbol", geneid$gene, mart=ensembl)
-    if (dim(genes.info)[1] == 0) return(NULL)
-    geneid %>%
-      left_join(genes.info, by=c('gene'='hgnc_symbol'))
-}
+
 
 #' Write csv file for data frame
 #'

### 2018-06-12 manuscript/main.tex
Table for vbsr analysis in manuscript

@@
 
-\subsection{Significant associated gene expression probes in \citet{figueroa2010dna}}
-
-Figure \ref{CC} shows the results of consensus clustering on probes. [Explain the choice of 7 clusters]
- $k = 7$ shows high intra-cluster consensus and low intercluster consensus.
-$k = 7$ approaches the maximum consensus distribution 
-$k = 7$ is the largest $k$ with a appreciable increase in consensus 
- and significantly enriched biological processes in each cluster is shown below in Table \ref{enriched}.
-
 \begin{table*}[ht!]
@@
   \midrule
-  cluster 1 (size 326) \\
+  Cluster 1 (size 326) \\
   \midrule
@@
    \midrule
-   cluster 3 (size 671) \\
+   Cluster 3 (size 671) \\
    \midrule
@@
 \midrule
-  cluster 5 (size 816)\\
+  Cluster 5 (size 816)\\
   \midrule
@@
      \midrule
-   cluster 6 (size 656)\\
+   Cluster 6 (size 656)\\
   \midrule
@@
      \midrule
-     cluster 7 (size 461) \\
+     Cluster 7 (size 461) \\
      \midrule
@@
 \end{tabular}
-\caption{\textbf{Enriched biological processes in \citet{figueroa2010dna} methylation probes cluster} }
+\caption{\textbf{Enriched biological processes in \citet{figueroa2010dna} methylation probes cluster} PANTHER overrepresentation test were performed using the associated gene with each methylation probe with control of $FDR < 0.05$.
+The table shows the test gene set (genes in each cluster), reference set size, overlap between the two sets, fold enrichment and corresponding raw p-value and FDR value.
+The hits with overlap between test and reference gene set at least 25\% of the cluster size were shown in the table.
+Cluster 2 (size 290) and 4 (size 525) result in no significant hits.}
 \label{enriched}
@@
 
-\subsection{vBsr}
+\subsection{Methylation probes clustering}
+
+Figure \ref{CC} shows the results of consensus clustering on probes.
+Consensus of a pair of probe is defined to be the proportion of clustering runs on the resampled dataset that two probes are clustered together.
+Consensus matrices for cluster count $k = 1, \dots, 10$ were examined and $k = 7$ shows high intra-cluster consensus and low intercluster consensus among all cluster counts.
+Figure \ref{CC}(a) shows heat map of consensus matrix for cluster count $k = 6, 7, 8$.
+The CDF plot Figure \ref{CC}(b) shows that $k = 7$ approaches the maximum consensus distribution.
+The Delta Area Plot (Figure \ref{CC}(c)) shows the area under the CDF curves and $k = 7$ is observed to have the largest $k$ with a appreciable increase in consensus.
+These evidences suggest a cluster count of 7.
+Figure \ref{CC}(d) shows the heat map of methylation level with 16 column cluster of patients and 7 row cluster of probes.
+
+We examine each clusters for enrichment of specific biological processes using PANTHER overrepresentation test \citep{mi2013large} and the annotations associated with each cluster were shown below in Table \ref{enriched}.
+Cluster 2 and 4 showed no significant enrichment results.
+
+\subsection{Selection of gene expression probes associated with methylation pattern}
+
+For each cluster, we averaged the methylation level across patients.
+Variational Bayes spike regression was used to select the gene expression probes that are significantly associated with the average methylation profile.
+Figure \ref{heatmap} shows the heat map for methylation level, average methylation profile, and selected gene expression level for cluster $1, 3, 5, 6, 7$.
+Table \ref{genes} showed the significantly associated gene probes with each cluster and corresponding annotation and coefficient from vBsr model.
 
 \begin{figure*}[htb!]
-\includegraphics[width=\textwidth]{Figures/vBsr}
-\caption{\textbf{Heatmap of methylation clusters, average methylation level and gene expression probes significantly associate with average methylation level}}
+\centering
+\includegraphics[width=0.8\textwidth]{Figures/vBsr}
+\caption{\textbf{Heatmap of methylation clusters, average methylation level and gene expression probes significantly associate with average methylation level} }
+\label{heatmap}
 \end{figure*}
+
+\begin{table*}[ht!]
+\centering
+\begin{tabular}{llr}
+  \toprule
+gene & description & coef \\ 
+  \midrule
+  Cluster 1 \\
+  \midrule
+ERBB2 & erb-b2 receptor tyrosine kinase 2 & 0.002 \\ 
+  ZBTB38 & zinc finger and BTB domain containing 38 & -0.00068 \\ 
+  HDAC2 & histone deacetylase 2 & 0.00059 \\ 
+  RBPJ & recombination signal binding protein for immunoglobulin kappa J region & -0.00019 \\ 
+  POU6F1 & POU class 6 homeobox 1 & 0.0023 \\ 
+  TAL1 & TAL bHLH transcription factor 1, erythroid differentiation factor & 0.0002 \\ 
+  CDCA7 & cell division cycle associated 7 & -0.00015 \\ 
+  AURKB & aurora kinase B & 0.0004 \\ 
+   \midrule
+   Cluster 3 \\
+   \midrule
+   MED13L & mediator complex subunit 13 like & -0.00015 \\ 
+  INTS6 & integrator complex subunit 6 & -0.00021 \\ 
+  MGMT & O-6-methylguanine-DNA methyltransferase & 0.0006 \\ 
+  SOX15 & SRY-box 15 & 0.00071 \\ 
+  TRAF7 & TNF receptor associated factor 7 & 0.00028 \\ 
+  ABHD14B & abhydrolase domain containing 14B & -0.00012 \\ 
+  PDLIM1 & PDZ and LIM domain 1 & 1.1e-05 \\ 
+  ZMYM4 & zinc finger MYM-type containing 4 & 0.00018 \\ 
+   \midrule
+   Cluster 5 \\
+   \midrule
+   TLR6 & toll like receptor 6 & 0.0007 \\ 
+  SSU72 & SSU72 homolog, RNA polymerase II CTD phosphatase & 0.00013 \\ 
+  BCL7A & BCL tumor suppressor 7A & 0.00021 \\ 
+  SARS & seryl-tRNA synthetase & -0.0003 \\ 
+  SFPQ & splicing factor proline and glutamine rich & -2.8e-05 \\ 
+  \midrule
+  Cluster 6 \\
+  \midrule
+  DEAF1 & DEAF1, transcription factor & 8.1e-05 \\ 
+  ATF7IP & activating transcription factor 7 interacting protein & -5.1e-05 \\ 
+  PSEN1 & presenilin 1 & -0.00022 \\ 
+  CREBZF & CREB/ATF bZIP transcription factor & 0.00031 \\ 
+  RBL2 & RB transcriptional corepressor like 2 & 0.00024 \\ 
+  VLDLR & very low density lipoprotein receptor & -6.4e-05 \\ 
+  ZC3H8 & zinc finger CCCH-type containing 8 & -0.00045 \\ 
+  ZBTB24 & zinc finger and BTB domain containing 24 & 0.00015 \\ 
+  \midrule
+  Cluster 7 \\
+  \midrule
+  MAMLD1 & mastermind like domain containing 1 & -0.00059 \\ 
+  AKT1 & AKT serine/threonine kinase 1 & 9e-05 \\ 
+  F2RL1 & F2R like trypsin receptor 1 & -0.00015 \\ 
+  GFI1 & growth factor independent 1 transcriptional repressor & 8.6e-05 \\ 
+  GLO1 & glyoxalase I & -3.8e-05 \\ 
+  HNMT & histamine N-methyltransferase & 0.00092 \\ 
+  MED1 & mediator complex subunit 1 & -0.00018 \\ 
+  PAK6 & p21 (RAC1) activated kinase 6 & 0.00033 \\ 
+  ZNF286A & zinc finger protein 286A & -0.004 \\ 
+  KIAA1549 & KIAA1549 & 0.00047 \\ 
+  ZBED5 & zinc finger BED-type containing 5 & -0.00034 \\ 
+  SCML1 & Scm polycomb group protein like 1 & -0.00043 \\ 
+  CRTC3 & CREB regulated transcription coactivator 3 & -8.9e-05 \\ 
+  SLC11A1 & solute carrier family 11 member 1 & 0.00019 \\ 
+  TP53BP1 & tumor protein p53 binding protein 1 & 0.00032 \\ 
+  CBX2 & chromobox 2 & -0.00037 \\ 
+  TRIM15 & tripartite motif containing 15 & 0.0005 \\ 
+   \bottomrule
+\end{tabular}
+\label{genes}
+\caption{\textbf{Gene expression probes significantly associated with methylation cluster}}
+\end{table*}
 

### 2018-06-12 manuscript/main.tex
Update the caption for module heat map and table for significantly associated gene expression probes

@@
 \includegraphics[width=0.8\textwidth]{Figures/vBsr}
-\caption{\textbf{Heatmap of methylation clusters, average methylation level and gene expression probes significantly associate with average methylation level} }
+\caption{\textbf{Heat map of methylation level, average methylation level and gene expression probes significantly associate with average methylation level} For methylation cluster 1, 3, 5, 6, 7, the methylation level is shown.
+Methylation level within a cluster is averaged across the patients and served as response variable for vBsr model.
+The gene expression level of probes significantly associated with the methylation level was standardized and displayed as heat map as well.
+}
 \label{heatmap}
@@
 \label{genes}
-\caption{\textbf{Gene expression probes significantly associated with methylation cluster}}
+\caption{\textbf{Gene expression probes significantly associated with methylation cluster} HGNC gene symbol, short description and vBsr coefficients of the genes significantly associated with the methylation pattern were shown in the table for cluster 1, 3, 5, 6, and 7.
+}
 \end{table*}

### 2018-06-14 manuscript/main.tex
Update method section and add todos

@@
 \usepackage{epstopdf}
+\usepackage{color}
 % hyperref makes hyperlinks in the resulting PDF.
@@
 DNA methylation and gene expression microarray data of 344 AML cases from \citet{figueroa2010dna} were downloaded from the GEO repository (GEO Accession number: GSE18700, GSE14468).
-Briefly, methylation data were obtained from custom design human promoter microarray for HELP assay, and gene expression data was obtained from Affymetrix Human Genome U133 Plus 2.0 Array.
+The methylation data were generated from custom design human promoter microarray for HELP assay and gene expression data was obtained from Affymetrix Human Genome U133 Plus 2.0 Array.
 
-Preprocessed gene expression data were obtained from \citet{gentles2015prognostic}, which normalized the raw CEL files with Affymetrix MAS5 algorithm and $\log_2$ transformed.
-We queried the GO terms on molecular function and biological process based on the HUGO gene symbol of the probes and filtered for probes that contain keywords related to transcription and methylation, which serves as predictors of our model.
+\subsubsection{Gene expression data preprocessing}
+Gene expression data were preprocessed by \citet{gentles2015prognostic}, which normalized the raw CEL files with Affymetrix MAS5 algorithm and performed $\log_2$ transformation.
+We filtered the gene expression probes based on the GO molecular function and biological process annotation.
+Probes with annotation related to transcription and methylation were kept as the predictors of our model.
 
-For methylation data analysis, we followed \citet{figueroa2010dna} and performed unsupervised hierarchical clustering of patients from methylation microarray using subset of the probes with standard deviation $> 1$ across all AML cases ($n = 3745$).
-Lingoes transformed 1 - Pearson correlation distance and Ward's minimum variance criterion were used.
+\subsubsection{Methylation data analysis}
+For methylation data, we followed \citet{figueroa2010dna} and filtered for probes with standard deviation $> 1$ across all AML cases ($n = 3745$).
+We performed hierarchical clustering on patients using Lingoes transformed 1 - Pearson correlation distance and Ward's method.
 The patient clusters from the original study was reproduced.
-Next, we clustered the probes with similar methylation pattern using consensus clustering \citep{monti2003consensus}, which produces visual and quantitative stability evidence to a given number of clusters $k$ and cluster assignments.
-R package \texttt{ConsensusClusterPlus} \citep{wilkerson2010consensusclusterplus} provides an implementation of the method and were used to determine the optimal number of clusters.
-We used the following options: 80 \% probe subsampling, Ward's criterion with Lingoes transformation on 1 - Pearson correlation distance, 50 replicates for each $k$  and maximum $k = 10$.
-We decides the number of optimal number of clusters based on the consensus matrix and area under CDF.
+
+We performed consensus clustering \citep{monti2003consensus} on the probes to determine the optimal number of probe clusters.
+R package \texttt{ConsensusClusterPlus} \citep{wilkerson2010consensusclusterplus} provides an implementation of consensus clustering and we used the following parameteres: 80 \% probe subsampling, Ward's criterion with Lingoes transformation on 1 - Pearson correlation distance, 50 replicates for each cluster number $k$  and maximum $k = 10$.
+We decides the number of optimal number of clusters based on the visual and quantitative evidence from the consensus matrix and area under CDF plot.
 The enrichment of biological processes represented by genes in each methylation cluster were examined using PANTHER overrepresentation test using Gene Ontology Biological process annotation data \citep{mi2013large} with $ FDR < 0.05$.
@@
 \subsection{Selection of significant gene expression probes}
-Variational Bayesian spike regression (vBsr) \cite{logsdon2012novel} is a penalized Bayesian regression model that imposes sparsity constraint on the regression coefficients using a spike-and-slab prior and utilized mean-field approximation to achieve fast computation.
-In addition, the algorithm was ran multiple times with random initialization to identify multiple local maxima of lower bound and provides the option of Bayesian Model Averaging (BMA) over the identified models to produce an estimate to reduce the model uncertainty.
-vBsr also defines a test statistic $z_{vb}$ associated with each penalized coefficients that allow control over the family-wise error rate by tuning the penalty parameter $l_0$ such that $z_{vb}$ statistics are approximately $\mathcal{N}(0, 1)$ under the null hypothesis.
-We used vBsr to select the subset of gene expression probes that significantly associate with the observed pattern of variation of methylation level across patients.
-We ran 100 random starts for each model and used BMA option.
-We tuned the penalty parameter $l_0$ such that a feature will have a posterior probability of 0.95 if it passes a Bonferroni correction in the multivariate model to control the Type I error rate.
+We used Variational Bayesian spike regression (vBsr) \cite{logsdon2012novel} to select gene expression probes that are significantly associated with methylation pattern across patients.
+vBsr is a penalized Bayesian regression model that uses a spike-and-slab prior to impose sparsity constraint on the regression coefficients.
+Fast computation were achieved by utilizing mean-field approximation.
+The algorithm was ran 100 times with random initialization to identify multiple local maxima of lower bound and we used the option of Bayesian Model Averaging (BMA) to produce a unique estimate over all identified models.
+vBsr defines a test statistic $z_{vb}$ associated with each penalized coefficients that allow control over the family-wise error rate by tuning the penalty parameter $l_0$  such that $z_{vb}$ statistics are approximately $\mathcal{N}(0, 1)$ under the null hypothesis.
+We tuned the penalty parameter such that a feature will have a posterior probability of 0.95 if it passes a Bonferroni correction in the multivariate model.
 Gene expression probes that were significant for $z_{vb}$ at $\widehat{FDR} = 0.1$ were selected.
@@
 \includegraphics[width=\textwidth]{Figures/clustering}
-\caption{\textbf{Consensus clustering of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$ (b) cumulative distribution function of consensus matrix at each $k = 1, \dots, 10$ (c) Area under CDF of consensus matrix for $k = 1, \dots, 10$ (d) Hierarchical clustering of cases and probes.
+\caption{\textbf{Consensus clustering of methylation probes for \citet{figueroa2010dna}} (a) consensus matrix for k = $6, 7, 8$ (b) cumulative distribution function of consensus matrix at cluster count $k \in \{1, \dots, 10\}$ (c) Area under CDF of consensus matrix for $k \in \{1, \dots, 10\}$ (d) Hierarchical clustering of patients and probes.
 Each row represents a probe and each column represents a patients.
@@
 The16 AML cases from \citet{figueroa2010dna} were reproduced.
-Probes were clustered using Ward's method with lingoes transformation of 1 - Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.}
+Probes were clustered using Ward's method with lingoes transformation of 1 - Pearson correlation distance transformed to Euclidean space and $k=7$ were chosen as the cutoff from CC.
+{\color{red} TODO: description of row and column normalizing procedure for heat map visualization}}
 \label{CC}
@@
 
-\subsection{TCGA LAML RNA-seq and DNA methylation data}
-TCGA Level 3 mRNAseq and DNA methylation data was downloaded from Broad TCGA GDAC site for LAML.
+\subsection{TCGA LAML RNA-Seq and DNA methylation data}
+
+TCGA LAML Level 3 mRNAseq and HumanMethylation450 BeadChip methylation data was downloaded from Broad TCGA GDAC site \citep{bitgda2016analysis}.
+
 mRNAseq probes with missing values were imputed by KNN using R packages \texttt{impute} \citep{hastie2001impute}.
-We queried the GO terms on molecular function and biological process based on the HUGO symbol of the probes and filtered for probes that contain keywords related to transcription and methylation as predictors for our model.
+We queried the GO terms on molecular function and biological process on the genes associated with the probes and filtered for probes that contain keywords related to transcription and methylation as predictors for our model.
 
-Methylation array data was obtained from Illumina Infinium HumanMethylation450 BeadChip.
-Probes that are on chromosome X and Y were removed.
+For methylation data, we removed probes that are on chromosome X and Y.
 Probes with UCSC RefGene group annotation as TSS 1500 and located within UCSC CpG island annotation were selected.
-The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with FDR adjusted p-value (FDR $< 0.1$) of 0.05.
-The resulting probes were clustered using hierarchical clustering with euclidean distance using Ward's method.
+{\color{red} Filtering method?} The correlation between the Beta value of filtered methylation probes and RSEM level of the corresponding genes in mRNAseq data were computed and significant methylation probes was filtered with $FDR < 0.1$.
+The resulting probes were clustered using hierarchical clustering with euclidean distance and Ward's method.
 20 clusters were set as the cutoff for assigning cluster membership.
@@
 \centering
+\begin{small}
 \begin{tabular}{lllllllll}
@@
 \end{tabular}
+\end{small}
 \caption{\textbf{Enriched biological processes in \citet{figueroa2010dna} methylation probes cluster} PANTHER overrepresentation test were performed using the associated gene with each methylation probe with control of $FDR < 0.05$.
@@
 Figure \ref{CC} shows the results of consensus clustering on probes.
-Consensus of a pair of probe is defined to be the proportion of clustering runs on the resampled dataset that two probes are clustered together.
+The consensus of a pair of probe is defined to be the proportion of clustering runs on the resampled dataset that two probes are clustered together \citep{monti2003consensus}.
 Consensus matrices for cluster count $k = 1, \dots, 10$ were examined and $k = 7$ shows high intra-cluster consensus and low intercluster consensus among all cluster counts.
@@
 \centering
+\begin{small}
 \begin{tabular}{llr}
@@
 \label{genes}
+\end{small}
 \caption{\textbf{Gene expression probes significantly associated with methylation cluster} HGNC gene symbol, short description and vBsr coefficients of the genes significantly associated with the methylation pattern were shown in the table for cluster 1, 3, 5, 6, and 7.
-}
+{\color{red} Use $\log_2$ (1 + TPM)}}
 \end{table*}

### 2018-07-26 R/model.R
Organize the folders

@@ -1,126 +1,0 @@
-# Author: Mamie Wang
-# Date created: 02/05/2018
-# Date last modified: 02/06/2018
-# R version: 3.4.2
-# Run lasso model for each methylation cluster using mRNA-seq genes as predictors
-#   - group methylation beta-value within each cluster
-#   - filter mRNA-seq for probes that corresponds to genes with keyword (methyl,
-#   transcription)
-
-library(glmnet)
-library(dplyr)
-
-load('../data/processed/clusterTCGA.RData')
-load('../data/processed/MethylationFiltered.RData')
-load('../data/processed/filteredGEP.RData')
-load('../data/processed/RNAseq.RData')
-load('../data/processed/transcriptionRelatedGenes.RData')
-
-# Convert beta values to M values
-Beta2M <- function(beta) {
-  log(beta/(1 - beta), base=2)
-}
-
-# add / subtract epsilon from 0 and 1
-AddSubtractEpsilon <- function(x) {
-  epsilon = .Machine$double.eps
-  if (x <= epsilon) x = x + epsilon
-  else if (1 - x <= epsilon) x = x - epsilon
-  x
-}
-M.filtered <- apply(M.filtered, 2, function(x) sapply(x, AddSubtractEpsilon))
-methylation.M <- apply(M.filtered, 2, function(x) sapply(x, Beta2M))
-
-# Group beta values within clusters
-averageMethylationProfile <- function(probeCluster, M) {
-    print('Average methylation level within clusters...')
-    M.grouped <- data.frame(M) %>%
-      dplyr::mutate(probe = as.character(rownames(M))) %>%
-      dplyr::left_join(probeCluster) %>%
-      dplyr::group_by(cluster) %>%
-      dplyr::select(-probe) %>%
-      dplyr::summarise_all(funs(mean)) %>%
-      na.omit()
-    M.grouped
-}
-
-distances <- dist(methylation.M) # euclidean distance
-hclust.M <- hclust(distances, method='ward.D2')
-methClusterNums = 20
-clusters <- data.frame(probe=rownames(methylation.M),
-                       cluster=cutree(hclust.M, methClusterNums))
-M.grouped <- averageMethylationProfile(clusters, methylation.M)
-save(M.grouped, file='../data/processed/averageMethylationProfile.RData')
-
-i = 1
[…]

### 2018-07-26 R/preprocessing.R

@@ -1,209 +1,0 @@
-# File name: preprocessing.R
-# Author: Mamie Wang
-# Date created: 02/05/2018
-# Date last modified: 02/05/2018
-# R version: 3.4.2
-# Preprocessing RNA-seq: 
-#   - remove NA (saved as data/processed/RNAseq.RData)
-#   - filter out low variance genes (sd < 0.5)
-#   - get annotation for probes and filter for those with terms 'transcrip' or 
-#     'methylation'
-# Preprocess methylation:
-#   - remove NA, chrX or chrY
-#   - control for age, gender
-#   - filter for low variance probes
-#   - hclust use Euclidean distance and Ward method
-
-
-library(dplyr)
-library(tidyr)
-library(mygene)
-library(ggplot2)
-
-
-methClusterNums = 20
-
-# Process RNA-seq data: remove NA
-print('Reading in mRNA-seq data...')
-RNAseq <- read.csv('../data/geneExpression/LAML.uncv2.mRNAseq_RSEM_normalized_log2.txt',
-                   sep='\t', header=T, stringsAsFactors=F)
-print('Preprocessing mRNA-seq data...')
-RNAseq <- na.omit(RNAseq)
-FormatPatientID <- function(x) substr(tolower(x), 1, 12)
-colnames(RNAseq) <- sapply(colnames(RNAseq), FormatPatientID)
-save(RNAseq, file='../data/processed/RNAseq.RData')
-
-# Process methylation data
-print('Reading in methylation data...')
-data.HM27 <- read.csv('../data/methylationHM27/LAML.methylation__humanmethylation27__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt', 
-                      sep='\t', header=T, stringsAsFactors=F)
-print('Preprocessing methylation data...')
-num.columns <- dim(data.HM27)[2]
-redundant.idx <- c(seq(7, num.columns, by=4), seq(8, num.columns, by=4), 
-                   seq(9, num.columns, by=4))
-data.HM27 <- data.HM27[-1,-redundant.idx]
-data.HM27[,seq(5)] <- data.HM27[,c(1,3,4,5,2)]
-columnnames <- colnames(data.HM27)
-columnnames[seq(4)] <- c('ID', 'Gene.Symbol', 'Chromosome', 'Genomic.Coordinate')
-colnames(data.HM27) <- columnnames
[…]

### 2018-07-26 R/visualization.R

@@ -1,52 +1,0 @@
-# Author: Mamie Wang
-# Date created: 02/06/2018
-# Date last modified: 02/06/2018
-# R version: 3.4.2
-# Module network visualization of the clusters
-
-library(ComplexHeatmap)
-library(viridis)
-
-ht_global_opt(heatmap_column_names_gp = gpar(fontsize = 6), 
-              heatmap_legend_title_gp= gpar(fontsize = 8),
-              heatmap_legend_labels_gp=gpar(fontsize = 6),
-              heatmap_row_names_gp = gpar(fontsize = 6))
-
-
-# average methylation level for each patient within each cluster
-load("../data/processed/averageMethylationProfile.RData")
-
-methClusterNums = 20
-i = 1
-for (i in seq(methClusterNums)) {
-  print(paste('Plotting for cluster', i))
-  # load Lasso model coefficients for cluster
-  load(paste0("../data/processed/coefs/coef", i, ".RData"))
-  
-  # load GEP level for the Lasso model nonzero coefficients for cluster
-  load(paste0("../data/processed/GEP/GEP", i, ".RData"))
-  
-  # load methylation level for cluster
-  load(paste0("../data/processed/Mcluster/Mcluster", i, ".RData"))
-  
-  M <- t(data.matrix(M.cluster[, seq(dim(M.cluster)[2]-2)]))
-  rownames(M) <- NULL
-  colnames(M) <- NULL
-  ht1 <- Heatmap(M, col=inferno(256), name='methylation', cluster_row=T, column_title='methylation')
-  
-  ha1 <- t(data.matrix(M.grouped[i, -1]))
-  rownames(ha1) <- NULL
-  colnames(ha1) <- NULL
-  ht2 <- Heatmap(ha1, col=inferno(256), width=unit(0.5, 'cm'), name='Average methylation')
-  
-  GEP <- scale(t(G.clusternonzero[as.character(nonzero.coefannot[,1]), ]),
-               center=TRUE, scale=FALSE)
-  rownames(GEP) <- NULL
-  colnames(GEP) <- NULL
-  ht3 <- Heatmap(GEP, col=inferno(256), name='GEP', column_title='GEP', cluster_columns=F)
-  
-  print(paste0('../data/figures/moduleVisualization/cluster', i, '.png'))
-  png(filename=paste0('../data/figures/moduleVisualization/cluster', i, '.png'))
-  draw(ht1 + ht2 + ht3)
-  dev.off()
-}

### 2018-08-07 analysis/Figueroa/methylation.tex
Update the methylation clustering analysis

\documentclass{article}
\usepackage{fullpage,graphicx}
\usepackage{amsmath,amsfonts,amsthm,amssymb,multirow,xcolor}
\usepackage{hyperref}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{pdfpages}
\usepackage[ruled,vlined,commentsnumbered,titlenotnumbered]{algorithm2e}
\usepackage[export]{adjustbox}

\begin{document}
\noindent
Methylation clustering analysis \hfill \textbf{Figueroa et al. 2010 data set} \newline 
\today \hfill Mamie Wang

\noindent
\rule{\linewidth}{0.4pt}

\listoffigures

\clearpage

\section{Preprocessing}

Methylation array data for 344 AML patients was downloaded from GEO \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18700}{GSE 18700} and 22759 probes that were mapped to a gene are filtered for downstream analysis (Each probe is annotated to the nearest transcription start site allowing for a maximum distance of 5 kb from the TSS). The distribution of the mean and standard deviation are shown below in Figure \ref{methylDistn}.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../../figures/Figueroa/methyl-distn}
\caption[Mean and standard deviation of the raw methylation data]{}
\label{methylDistn}
\end{center}
\end{figure}

Successive normalization were performed on the methylation array \cite{olshen2010successive} until the absolute difference between Frobenius norm of consecutive iterations is below $10^{-8}$. PhenoGraph were applied on the filtered on the normalized dataset \cite{levine2015data} which constructs a nearest-neighbor graph ($k = 30$) with weight between each probe pair as Jaccard similarity coefficients between the corresponding nearest neighbor set and performs Louvain community detection algorithm by maximizing the modularity on the graph. 20 communities (modules) were obtained as a result. The ``eigengene'' profile was computed for each module as their first principal component, shown in the Figure \ref{eigenGene}.

\begin{figure}[htbp]
\begin{center}

\caption[Eigengene methylation profile of the modules]{Eigengene methylation profile of the modules. Each black line represents a methylation level of a probe across patients. The blue line shows the eigengene expression profile.}
\label{eigenGene}
\end{center}
\end{figure}

\clearpage

\appendix

\clearpage

\end{document}

### 2018-09-03 analysis/Figueroa/methylation.tex
Complete eigen profile plots

@@
 
-Methylation array data for 344 AML patients was downloaded from GEO \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18700}{GSE 18700} and 22759 probes that were mapped to a gene are filtered for downstream analysis (Each probe is annotated to the nearest transcription start site allowing for a maximum distance of 5 kb from the TSS).
-The distribution of the mean and standard deviation are shown below in Figure \ref{methylDistn}.
+\subsection{Gene expression array data}
+
+Gene expression array data for 344 AML patients was downloaded from GEO \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE14468}{(GSE 14468)}.
+Normalization was performed in \cite{gentles2015prognostic}.
+Each probe was annotated to the gene with nearest transcription start site and the corresponding gene symbol was provided in the file.
+The gene symbols of the probes are used to query the Gene Ontology \cite{ashburner2000gene} for biological process annotation.
+We are interested in genes related to methylation and transcriptional regulation.
+To remove irrelevant genes, we filter for genes that have keyword ``methylation'' and ``transcrip'' in their biological process annotation.
+In the end, 3417 out of 17788 remained for the downstream analysis.
+
+\subsection{Methylation array data}
+
+Methylation array data for 344 AML patients was downloaded from GEO \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18700}{(GSE 18700)}.
+Each probe was annotated to the nearest transcription start site allowing for a maximum distance of 5 kb from the TSS.
+Any probes that were mapped to a gene are filtered for downstream analysis and in total 22759 out of 25626 probes are left after the filtering step.
+The distribution of the mean and standard deviation of the probes across patients are shown below in Figure \ref{methylDistn}.
 
@@
 Successive normalization were performed on the methylation array \cite{olshen2010successive} until the absolute difference between Frobenius norm of consecutive iterations is below $10^{-8}$.
-PhenoGraph were applied on the filtered on the normalized dataset \cite{levine2015data} which constructs a nearest-neighbor graph ($k = 30$) with weight between each probe pair as Jaccard similarity coefficients between the corresponding nearest neighbor set and performs Louvain community detection algorithm by maximizing the modularity on the graph.
+PhenoGraph  \cite{levine2015data} were used to cluster probes.
+PhenoGraph first constructs a nearest-neighbor graph ($k = 30$) which has individual probe as a node and edge weight between each probe pair as Jaccard similarity coefficients between their nearest neighbor set.
+Louvain community detection algorithm was performed to find the clustering assignment that maximizes the modularity on the graph.
 20 communities (modules) were obtained as a result.

### 2018-09-05 analysis/Figueroa/methylation.tex
Update the methylation clustering analysis

@@
 \usepackage[export]{adjustbox}
-
+\usepackage[margin=0.5in]{geometry}
 \begin{document}
@@
 Any probes that were mapped to a gene are filtered for downstream analysis and in total 22759 out of 25626 probes are left after the filtering step.
-The distribution of the mean and standard deviation of the probes across patients are shown below in Figure \ref{methylDistn}.
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../../figures/Figueroa/methyl-distn}
-\caption[Mean and standard deviation of the raw methylation data]{}
-\label{methylDistn}
-\end{center}
-\end{figure}
-
-Successive normalization were performed on the methylation array \cite{olshen2010successive} until the absolute difference between Frobenius norm of consecutive iterations is below $10^{-8}$.
+Normalization was performed on the probes across patients.
+%Successive normalization were performed on the methylation array \cite{olshen2010successive} until the absolute difference between Frobenius norm of consecutive iterations is below $10^{-8}$.
 PhenoGraph  \cite{levine2015data} were used to cluster probes.
@@
 20 communities (modules) were obtained as a result.
-The ``eigengene'' profile was computed for each module as their first principal component, shown in the Figure \ref{eigenGene}.
+The counts of probes in each cluster along with the ``eigengene'' profile was computed for each module as their first principal component, shown in the Figure \ref{eigenGene}.
 
@@
 \begin{center}
+\begin{tabular}{@{}c@{\hspace{.0cm}}@{\hspace{.0cm}}c@{\hspace{.0cm}}c@{\hspace{.0cm}}c@{}}
+\includegraphics[width=0.25\textwidth, page=1]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=2]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=3]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=4]{../../figures/Figueroa/eigen-profile} \\
+\includegraphics[width=0.25\textwidth, page=5]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=6]{../../figures/Figueroa/eigen-profile} & 
+\includegraphics[width=0.25\textwidth, page=7]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=8]{../../figures/Figueroa/eigen-profile} \\
+\includegraphics[width=0.25\textwidth, page=9]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=10]{../../figures/Figueroa/eigen-profile} & 
+\includegraphics[width=0.25\textwidth, page=11]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=12]{../../figures/Figueroa/eigen-profile} \\
+\includegraphics[width=0.25\textwidth, page=13]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=14]{../../figures/Figueroa/eigen-profile} & 
+\includegraphics[width=0.25\textwidth, page=15]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=16]{../../figures/Figueroa/eigen-profile} \\
+\includegraphics[width=0.25\textwidth, page=17]{../../figures/Figueroa/eigen-profile} &
+\includegraphics[width=0.25\textwidth, page=18]{../../figures/Figueroa/eigen-profile} \\
+\end{tabular}
+\begin{small}
+\begin{tabular}{rrrrrrrrrrrrrrrrrrr}
+  \hline
+module & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 \\ 
+  \hline
+probes & 3312 & 2810 & 2140 & 2013 & 1853 & 1842 & 1583 & 1325 & 1085 & 995 & 861 & 723 & 723 & 630 & 326 & 277 & 237 &  24 \\ 
+   \hline
+\end{tabular}
+\end{small}
+\caption[First principal component of methylation modules]{\textbf{First principal component of methylation modules} Each panel represent a module and the black lines represent the normalized methylation level of a probe across 344 patients.
+The blue lines show the first principal component of each module.
+The size of each module is shown in the table}
+\label{eigenGene}
+\end{center}
+\end{figure}
 
-\caption[Eigengene methylation profile of the modules]{Eigengene methylation profile of the modules.
-Each black line represents a methylation level of a probe across patients.
-The blue line shows the eigengene expression profile.}
-\label{eigenGene}
+In addition, the heatmap of expression level of the clusters are shown in Appendix Figure \ref{heatmap} and annotated by hits on GSEA.
+
+\clearpage
+
+\section{Variational Bayes Spike Regression}
+
+\clearpage
+
+\appendix
+
+\section{Methylation array heatmap}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{../../figures/Figueroa/methyl-clustering}
+\caption[Heatmap of methylation modules]{\textbf{Heatmap of methylation modules} Heatmap of methylation array expression.
+The rows are annotated with the PhenoGraph clusters and columns are arrange by hierarchical clustering on patients using Euclidean distance.
+}
+\label{heatmap}
 \end{center}
@@
 
-\appendix
-
-\clearpage
-
 \end{document}

### 2018-09-05 analysis/Figueroa/methylation.tex
Description for the methylation clustering

@@
 Gene expression array data for 344 AML patients was downloaded from GEO \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE14468}{(GSE 14468)}.
-Normalization was performed in \cite{gentles2015prognostic}.
-Each probe was annotated to the gene with nearest transcription start site and the corresponding gene symbol was provided in the file.
-The gene symbols of the probes are used to query the Gene Ontology \cite{ashburner2000gene} for biological process annotation.
-We are interested in genes related to methylation and transcriptional regulation.
-To remove irrelevant genes, we filter for genes that have keyword ``methylation'' and ``transcrip'' in their biological process annotation.
-In the end, 3417 out of 17788 remained for the downstream analysis.
+Normalization was performed in Gentles et al 2015 \cite{gentles2015prognostic}.
+Annotation of the probes to gene with nearest transcription start site were available from GEO.
+We query for the Gene Ontology terms of the corresponding genes and filtered the probes based on their GO biological process annotation to obtain genes related to methylation and transcriptional regulation.
+After filter for genes that have keyword ``methylation'' or ``transcrip'', 3417 out of 17788 are left for the downstream analysis.
 
@@
 Methylation array data for 344 AML patients was downloaded from GEO \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE18700}{(GSE 18700)}.
-Each probe was annotated to the nearest transcription start site allowing for a maximum distance of 5 kb from the TSS.
-Any probes that were mapped to a gene are filtered for downstream analysis and in total 22759 out of 25626 probes are left after the filtering step.
+Annotation of the probes to the nearest transcription start site allowing for a maximum distance of 5 kb from the TSS was provided along with the dataset.
+Any probes that were mapped to a gene are filtered for downstream analysis and in total 22759 out of 25626 probes are left after this step.
 
-Normalization was performed on the probes across patients.
+Standardization was performed on the probes across patients.
 %Successive normalization were performed on the methylation array \cite{olshen2010successive} until the absolute difference between Frobenius norm of consecutive iterations is below $10^{-8}$.
 PhenoGraph \cite{levine2015data} were used to cluster probes.
-PhenoGraph first constructs a nearest-neighbor graph ($k = 30$) which has individual probe as a node and edge weight between each probe pair as Jaccard similarity coefficients between their nearest neighbor set.
-Louvain community detection algorithm was performed to find the clustering assignment that maximizes the modularity on the graph.
-20 communities (modules) were obtained as a result.
-The counts of probes in each cluster along with the ``eigengene'' profile was computed for each module as their first principal component, shown in the Figure \ref{eigenGene}.
+PhenoGraph first constructs a nearest-neighbor graph ($k = 30$) that has individual probe as a node.
+The edge weight between probe pairs are computed as the Jaccard similarity coefficients between their nearest neighbor set.
+Louvain community detection was performed in this algorithm to find the clustering assignment that maximizes the modularity on the graph.
+After applying PhenoGraph on the 22759 methylation probes, 20 communities (modules) were obtained.
+The counts of probes in each module along with the first principal component of each module are shown in the Figure \ref{eigenGene}.
 
@@
 
-In addition, the heatmap of expression level of the clusters are shown in Appendix Figure \ref{heatmap} and annotated by hits on GSEA.
-
-\clearpage
-
-\section{Variational Bayes Spike Regression}
+In addition, the heatmap of expression level of the clusters are shown in Appendix Figure \ref{heatmap}.
 
@@
 
+\bibliography{methylation} 
+\bibliographystyle{ieeetr}
+
 \end{document}

### 2018-09-10 analysis/Figueroa/methylation.tex
Update vBsr plot

@@
 
+\subsection{Selection of significant gene expression probes using vBsr}
+We used Variational Bayesian spike regression (vBsr) \cite{logsdon2012novel} to select gene expression probes that are significantly associated with module methylation profile.
+vBsr is a penalized Bayesian regression model that uses a spike-and-slab prior to impose sparsity constraint on the regression coefficients.
+Fast computation were achieved by utilizing mean-field approximation.
+The algorithm was ran 50 times with random initialization to identify multiple local maxima of lower bound and Bayesian Model Averaging (BMA) was used to produce a unique estimate over all identified models.
+vBsr defines a test statistic $z_{vb}$ and corresponding p-value.
+We applied Bonferroni correction to the p-values at $\alpha = 0.05$.
+Figure \ref{vbsr} shows the $-\log p$ values of the gene probes for modules that have at least one significant hits.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../../figures/Figueroa/vbsr.png}
+\caption[Gene expression probes significantly associated with methylation module profile selected by vBsr]{\textbf{Gene expression probes significantly associated with methylation module profile selected by vBsr} The Bonferroni corrected significant level is shown as the dashed lines and the significant hits are annotated with corresponding gene symbol.
+}
+\label{vbsr}
+\end{center}
+\end{figure}
+
+\clearpage
+
 \appendix
